Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-23 paper/main.tex
adapt model text; clear some of christian thoughts; add table with parameters and variables

We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness of the bacterial rod~\cite{Amir2014_2}.
As explained above, MReB will counteract the bending of the rod, resulting in a stiffening force.
We model the bacterium as a bending beam.
% In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988},
% but for sake of simplicity we chose to model the bacterium as a bending beam.
% \textbf{[", thus utilizing the simplest solution which can still produce the desired effects." Es ist unklar was die "desired effects" sind. ]}\\
Using our discretization $\vec{x}_i$, we connect the vertices by springs of length $l_i$.
% which can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
% springs of lengths $l_i$.
% \textbf{[Das klingt nach einer Wiederholung von Spatial Representation. Du erklärst ziemlich umständlich. Konzentriere dich auf das wesentliche und überlege, was du eigentlich sagen möchtest.]}\\
We use the notation $\vec{w}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

@@

The strength of the potential is given by $\epsilon$ and the thickness of our rods
is encoded in $\sigma = (m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the thicknesses of the two
interacting bacteria.
We utilize the shorthand notation $r=|\vec{x}_i - \vec{p}|$ for the distance between the two points
in question.
We numerically regularize the potential by imposing an upper bound $\beta$ on the norm of the
resulting force $\vec{F} = \vec{\nabla} V$.
Furthermore, we impose a cutoff parameter $\zeta$ which sets sets the force to zero when the
distance $r$ between the interacting points exceeds this threshold.
The resulting force is given by

@@

The physical interaction between the bacteria plays a crucial role for the overall dynamics of our
simulation.
The Mie potential is capable of describing a wide range of interaction types from soft to rigid, by
adjusting the parameters of the potential accordingly.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower amount of parameters and more restrictive mathematical formulation.
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low amount of parameters that can be estimated by a data-driven approach.

@@

To simulate proliferation, we introduce a growth term for the spring lengths $l_i$.
In most of our simulations, we choose to use identical lengths for all segments (i.e. $l_i=l$).
However, in order to generalize our approach we assume that individual lengths of segments could
vary. Our implemented model allows us to reduce the growth rate with increasing number of neighbours
$n\leq N$. This option provides a phenomenological approach for modelling slower growth in larger colonies but
is be toggled off by default thus removing the additional term in brackets of equation~\ref{eq:growth-rod-lengths}.

\textbf{[Ich empfinde deinen Text als uneinheitlich. Du solltest hier nicht deine Softwareimplementation beschreiben, sondern das was du tatsächlich in den Simulationen benutzt hast. Also Möglichkeiten, die du in deinem Code angelegt hast, aber für das Paper nicht verwendet hast, brauchst du nicht zu erwähnen. Das führt zu Verwirrung.]}

\begin{equation}
    \partial_t l_i = \mu\left(1-\frac{\text{min}(n,N)}{N}\right)l_i
    \label{eq:growth-rod-lengths}
\end{equation}

Note that the growth rate is proportional to the rod length $l$ which leads to exponential
growth~\cite{Robert2014,Takeuchi2005}.
\textbf{[Das zu erwähnen ist eine Trivialität für ein Forschungspapier. Du kannst begründen warum du exp. Wachstum verwendest und wie du den Teilungszeitpunkt bestimmst.]}

This process will increase the length of the cell indefinitely unless we impose a condition for the
division event. We define a fixed length threshold for the total length of the cell at which it divides.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.
We first calculate the new spring lengths of the two resulting rods, according to the equation:

\begin{equation}
    \tilde{l} = l_i\left(\frac{1}{2} - \frac{r}{\sum\limits_i l_i}\right)
    \label{eq:growth-rod-new-lengths}
\end{equation}

Afterwards, we start at one endpoint $\vec{x}_0$ and move along the polygon defined by the
bacterial rod. We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{y}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.
\textbf{[Was ist $r$? Was ist mit den Parametern und mit anderen intro-zellularen Zuständen?]}

@@

We implemented the described model in our recently developed \ac{abm} framework
\texttt{cellular\_raza}~\cite{Pleyer2025}.
By design \texttt{cellular\_raza} does not assume any particular cellular representation, thus makes
it flexible and easy to adapt the simulation to particular needs. This allows us to experiment with
different model implementations and still retain performance.
\textbf{[Der Zusammenhang zwischen der Flexibilität und Performance ist unklar.]}

Furthermore, it is an \ac{abm} which is able to natively track cells through the course of a
numerical simulation and also keeps track of mother/daughter relationships.
\textbf{[Das scheint hier trivial zu sein.]}

On the programming side, we are using the \texttt{pyo3} and \texttt{maturin} crates which allow us
to easily generate Python bindings and enable us to interoperate between the two ecosystems.
\textbf{[Das kennt hier keiner und gehört eher in den Anhang.]}

By using \texttt{cellular\_raza}, we can make use of multiple threads for running a single numerical
simulation, using the built-in parallelization methods (see Supplementary Material,
Figure~\ref{fig:performance}).
\textbf{[Ist alles eher für den Anhang. Vielleicht solltest du mehr darüber reden, warum ABMs für die Fragestellungen gut geeignet sind.]}

@@

To remove this non-identifiability we reduce the number of parameters by setting $\gamma=\gamma_0/s$
and $\eta=\eta_0 s$ for the rod rigidity and spring tension, resp.
We then optimize for this parameter and pick $\gamma_0=\SI{15}{\pixel}$ with
$\eta_0=\SI{150}{\pixel\per\second\squared}$ which corresponds to $60\%$ of the previously considered
interval. \textbf{[Unklar, wie die Pixel hier auftauchen.]}
The corresponding plots can be seen in \ref{subsec:supplement-parameter-profiles-amir}.

@@

%###################################################################################################
\subsection{Performance (TODO)}
\label{section:supplement-performance}

>>>>

We use principles from discrete differential geometry~\cite{bobenko2008discrete} to describe the
stiffness of the bacterial rod~\cite{Amir2014_2}.
As explained above, MReB will counteract the bending of the rod, resulting in a stiffening force.
We model the bacterium as a bending beam.
% In principle, this stress will be dependent on the position inside the cell~\cite{Chatterjee1988},
% but for sake of simplicity we chose to model the bacterium as a bending beam.
% \textbf{[", thus utilizing the simplest solution which can still produce the desired effects." Es ist unklar was die "desired effects" sind. ]}\\
Using our discretization $\vec{x}_i$, we connect the vertices by springs of length $l$.
% which can be represented by a collection of vertices $\vec{x}_i$ which are connected by a
% springs of lengths $l$.
% \textbf{[Das klingt nach einer Wiederholung von Spatial Representation. Du erklärst ziemlich umständlich. Konzentriere dich auf das wesentliche und überlege, was du eigentlich sagen möchtest.]}\\
We use the notation $\vec{w}_i:=\vec{x}_i-\vec{x}_{i-1}$ with which the resulting force acting
between two vertices if given by

@@

The strength of the potential is given by $\epsilon$ and the thickness of our rods
is encoded in $\sigma = (m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the thicknesses (radii) of the
two interacting rod-like bacteria.
We utilize the shorthand notation $r=|\vec{x}_i - \vec{p}|$ for the distance between the two points
in question.
We numerically regularize the potential by imposing an upper bound $\beta$ on the norm of the
resulting force $\vec{F} = \vec{\nabla} V$.
Furthermore, we impose a cutoff parameter $\zeta$ which sets sets the force to zero when the
distance $r$ between the interacting points exceeds this threshold.
The resulting force is given by

@@

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
By varying its parameters, the Mie potential is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels that correspond to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower amount of parameters and more restrictive mathematical formulation.
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low amount of parameters that can be estimated by our data-driven approach.

@@

To simulate cellular growth, we introduce an exponential growth term for the spring lengths $l$
\cite{Takeuchi2005}.

\begin{equation}
    \partial_t l = \mu l
\end{equation}

This process will increase the length of the cell indefinitely unless we impose a condition for the
division event.
The cell divides when the spring length $l$ has reached a fixed length threshold~\cite{Robert2014}
$l_\text{thresh}$.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.
We first calculate the new spring lengths of the two resulting rods via

\begin{equation}
    \tilde{l} = l\left(\frac{1}{2} - \frac{R}{n_v l}\right)
\end{equation}

where $n_v$ is the number of vertices and $R$ the previously introduced thickness of the rod.
This relation ensures that that the tips of the two new rods are touching right after the division
event.
We subdivide the polygon given by the vertices of the mother cell in $n$ segments of length
$\tilde{l}$, one segment of length $2R$ and again $n$ segments of length $\tilde{l}$.
The first and latter groups determine the new vertices of the two daughter cells.
This approach assumes cells where the total length of the polygon is greater than its diameter
$n_vl>2R$.
We set the growth rates of the daughter cells explicitly during the division event.
This approach allows us to treat the growth rates of the daughter cells as full individual
parameters within our model.
In order to reuse this model, we also provide the option to assign growth rates by sampling from a
normal distribution with user-defined parameters.
Every other intracellular parameter is copied identically to the daughter cells.
Table~\ref{table:list-of-parameters} shows a list of all parameters used within our model.

\begin{table}
    \centering
    \def\arraystretch{1.3}
    \begin{tabular}{l c l c}
        \textbf{Parameters} && \textbf{Variables}\\
        \toprule
        \multicolumn{4}{l}{\textbf{Mechanics}}\\
        \midrule
        Spring tension & $\gamma$ & Position (vertices) & $\vec{x}_i$\\
        Rigidity & $\eta$ & Velocity & $\partial_t\vec{x}_i$\\
        Damping & $\lambda$ & Spring length & $l$ \\
        \multicolumn{4}{l}{\textbf{Interaction}}\\
        \midrule
        Thickness & $R$\\
        Strength & $\epsilon$\\
        Exponent 1 & $n$\\
        Exponent 2 & $m$\\
        Bound & $\beta$\\
        Cutoff & $\zeta$\\
        \multicolumn{4}{l}{\textbf{Growth \& Division}}\\
        \midrule
        Growth rate & $\mu$\\
        Spring length threshold & $l_\text{thresh}$\\
        \bottomrule
    \end{tabular}
    \caption{
        List of all parameters and variables.
    }
    \label{table:list-of-parameters}
\end{table}

@@

We implemented the described model in our recently developed \ac{abm} framework
\texttt{cellular\_raza}~\cite{Pleyer2025}.
By design \texttt{cellular\_raza} does not assume any particular cellular representation, thus makes
it flexible and easy to adapt the simulation to particular needs.
Its design allows us to experiment with different model implementations while retaining fast
execution times.
\textbf{[Der Zusammenhang zwischen der Flexibilität und Performance ist unklar.]}
\textit{Oft muss man das eine für das andere opfern. Hier nicht/kaum.}

Furthermore, it is an \ac{abm} which is able to natively track cells through the course of a
numerical simulation and also keeps track of mother/daughter relationships.
\textbf{[Das scheint hier trivial zu sein.]}
\textit{Machen aber viele (zB. PhysiCell) einfach nicht. (Mother-Daughter relation)}

@@

To remove this non-identifiability we reduce the number of parameters by setting $\gamma=\gamma_0/s$
and $\eta=\eta_0 s$ for the rod rigidity and spring tension, resp.
We then optimize for this parameter and pick $\gamma_0=\SI{15}{\micro\metre}$ with
$\eta_0=\SI{150}{\micro\metre\per\second\squared}$ which corresponds to $60\%$ of the previously considered
interval.
The corresponding plots can be seen in \ref{subsec:supplement-parameter-profiles-amir}.

@@

%###################################################################################################
\subsection{Implementation \& Performance (TODO)}
\label{section:supplement-performance}

On the programming side, we are using the \texttt{pyo3} and \texttt{maturin} crates which allow us
to easily generate Python bindings and enable us to interoperate between the two ecosystems.

By using \texttt{cellular\_raza}, we can make use of multiple threads for running a single numerical
simulation, using the built-in parallelization methods (see Supplementary Material,
Figure~\ref{fig:performance}).
